Polyester RNA-seq simulation study

### 2018-07-23 datasets/RNA-Seq/polyester-simulation.Rmd
Add flexibility to simulate arbitrary number of genes

```{r workspace-setup, results='hide', message=FALSE, warning=FALSE}
# Load packages needed to simulate RNA-seq counts with polyester
library(SummarizedBenchmark)
library(zebrafishRNASeq) 
library(genefilter)
library(limma)
library(polyester)
library(DESeq2)
library(data.table)
library(magrittr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(tibble)
library(ggthemes)
library(readr)

@@

# results files that will be generated below
resfile_n5 <- file.path(resdir, "polyesterY-results-null5.rds")
resfile_n10 <- file.path(resdir, "polyesterY-results-null10.rds")
resfile_d5 <- file.path(resdir, "polyesterY-results-de5.rds")
resfile_d10 <- file.path(resdir, "polyesterY-results-de10.rds")

resfile_n5_uninfCov <- file.path(resdir, "polyesterY-results-null5-uninfCov.rds")
resfile_n10_uninfCov <- file.path(resdir, "polyesterY-results-null10-uninfCov.rds")
resfile_d5_uninfCov <- file.path(resdir, "polyesterY-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "polyesterY-results-de10-uninfCov.rds")

@@

We use the [polyester](https://bioconductor.org/packages/release/bioc/html/polyester.html) Bioconductor package to simulate RNA-Seq reads. The package requires a count matrix as input which estimates the mean variance relationship. Here we start with the [zebrafish](https://bioconductor.org/packages/release/data/experiment/html/zebrafishRNASeq.html) Bioconductor data package.

Estimate mean and variance relationship

```{r estimate-parameters-NB}
data(zfGenes)
dim(zfGenes)

filter <- apply(zfGenes, 1, function(x) length(x[x>5])>=2)
counts <- as.matrix(zfGenes[filter,])

Try using the yeast data instead - want all samples of the same condition, and the
zebrafish dataset only has 3v3.

@@

```{r}
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param rseed random seed
simulateOneSplit <- function(X, rseed, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE){
  if(uninformativeCovariate)
    uninf_covar <- rnorm(nrow(counts))

@@

  # NULL simulation
  if(nDE == 0) {
    truth <- rep(FALSE, nrow(counts)) 
    names(truth) <- rownames(counts)

    # log2FC are all 0 in null simulation
    log2FC = rep(0,nrow(counts))
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

  # DE simulation
  if(nDE > 0){
    # pick random set of nDE genes to add signal to
    DE <- sample(1:nrow(counts), nDE)
    truth <- rep(FALSE, nrow(counts))
    truth[DE] <- TRUE
    names(truth) <- rownames(counts)

    # randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
    log2FC <- rep(0, nrow(counts))
    log2FC[DE] <- rnorm(nDE, sd = 1)
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

@@

  geneExp <- tbl_df(data.frame(pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = res_alt$baseMean, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth, 
                               pzero = rowSums(counts(dds_alt)==0)/
                                 ncol(counts(dds_alt))  ))

@@

  # filter NAs and those with less than 50% expressed 
  geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

  # built Benchmark data
  sb <- bd %>% buildBench(data=geneExp, parallel = FALSE, 
                          truthCols = "qvalue", 
                          ftCols = "ind_covariate")
  rowData(sb)$log2FC <- geneExp$effect_size

>>>>

```{r workspace-setup, results='hide', message=FALSE, warning=FALSE}
# Load packages needed to simulate RNA-seq counts with polyester
library(SummarizedBenchmark)
library(genefilter)
library(limma)
library(polyester)
library(DESeq2)
library(data.table)
library(magrittr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(tibble)
library(ggthemes)
library(readr)

@@

# results files that will be generated below
resfile_n5 <- file.path(resdir, "polyester-results-null5.rds")
resfile_n10 <- file.path(resdir, "polyester-results-null10.rds")
resfile_d5 <- file.path(resdir, "polyester-results-de5.rds")
resfile_d10 <- file.path(resdir, "polyester-results-de10.rds")

resfile_n5_uninfCov <- file.path(resdir, "polyester-results-null5-uninfCov.rds")
resfile_n10_uninfCov <- file.path(resdir, "polyester-results-null10-uninfCov.rds")
resfile_d5_uninfCov <- file.path(resdir, "polyester-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "polyester-results-de10-uninfCov.rds")

@@

We use the [polyester](https://bioconductor.org/packages/release/bioc/html/polyester.html) Bioconductor package to simulate RNA-Seq reads. The package requires a count matrix as input which estimates the mean variance relationship. Here we start with the yeast data with 48 biological replicates in each of 
two conditions (analyzed in this [publication](https://www.ncbi.nlm.nih.gov/pubmed/26206307/)). 
We'll use the samples passing QC in the original study that belong to the WT group to 
estimate mean and variance relationship. We assume the data has already been downloaded
(this is carried out in the `yeast-simulation.Rmd` vignette).

@@

```{r}
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param rseed random seed
simulateOneSplit <- function(X, rseed, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE){
  ngenes <- 10000

  uninf_covar <- rnorm(ngenes)

@@

  gs <- sample(1:nrow(counts), ngenes, replace = TRUE)
  mu <- paramPolyester$mu[gs]
  p0 <- paramPolyester$p0[gs]

  # Null genes
  log2FC = rep(0, ngenes)
  dat_alt = create_read_numbers(mu,
                                paramPolyester$fit,
                                p0,
                                n=sampleSize*2, beta=cbind(log2FC), 
                                mod=modPolyester)

  # filter genes with mostly zeroes
  pzero = rowSums(dat_alt==0)/ncol(dat_alt)
  dat_alt <- dat_alt[pzero < 0.5,]
  truth <- rep(FALSE, nrow(dat_alt))
  log2FC <- log2FC[pzero < 0.5]
  mu <- mu[pzero < 0.5]
  p0 <- p0[pzero < 0.5]
  uninf_covar <- uninf_covar[pzero < 0.5]

  # DE simulation
  if(nDE > 0){
    # pick random set of nDE genes to add signal to
    DE <- sample(1:nrow(dat_alt), nDE)
    truth[DE] <- TRUE

    # randomly sample a log2FC from N(0,1) 
    log2FC[DE] <- rnorm(nDE, sd = 1)
    dat_alt[DE,] = create_read_numbers(mu[DE], paramPolyester$fit, p0[DE], 
                                  n=sampleSize*2, beta=cbind(log2FC[DE]), 
                                  mod=modPolyester)
  }

@@

  geneExp <- tbl_df(data.frame(pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = res_alt$baseMean, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth))

@@

  # filter NAs 
  geneExp <- geneExp %>% na.omit()

  # built Benchmark data
  sb <- bd %>% buildBench(data=geneExp, parallel = FALSE, 
                          truthCols = "qvalue", 
                          ftCols = "ind_covariate")
  sb <- addDefaultMetrics(sb)
  rowData(sb)$log2FC <- geneExp$effect_size
